SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-20 paper.tex
revamped abstract

\begin{abstract}
Short stretches of amino acid residues called targeting signals are involved in the proper localization of proteins in a cell, which is essential to maintain their desired function. Ones of them are signal peptides, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction. Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability. The proposed approach is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. By using n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides. Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor despite the fact it was trained on very small data set. Moreover, it appears to be very stable regardless of types of learning data sets. Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized that our model described signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84). The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.

>>>>

\begin{abstract}
Signal peptides are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. They may be recognized with adequate accuracy using learning systems, that are usually 'black-box' models. Without an ability to trace decision rules, researchers cannot discover which parameters are responsible for predictions. Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation. The proposed approach is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. Using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides. Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even after training on very small data sets. Moreover, it appeared very stable regardless of types of learning data sets. Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized that our model described signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84). The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.
